---
title: "Diabetes prediction using regression analysis in R"
#author: "Melvin Galera"
#date: "24-03-2024"
output: 
  github_document:
    toc: false
---

```{r setup, include=FALSE}
# Global code chunk options; adjust individual codes as required
knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE, fig.width = 12, fig.height = 12)

# Load data
diabetes_df <- read.csv("./data/diabetes.csv")

# Load libraries
library(tidyverse)
library(cowplot)
library(kableExtra)
library(scales)
library(ggcorrplot)
library(gridExtra)
library(ggpubr)
library(Amelia)
library(sjPlot)
library(arm)
library(caret)
library(pROC)

```

<br>

### I. Project objective

Diabetes is a serious chronic condition where there is too much glucose in the blood. Blood glucose is regulated by the hormone insulin. In such condition, the body cannot produce enough insulin or is not effectively using the insulin it makes. This can lead to excess sugar in the blood, which over time, can cause damage to blood vessels and nerves, and may result to many health complications. Early diagnosis leading to early and optimal self-management and treatment of diabetes is important to reduce risks of diabetes-related complications.

In this regard, this project was aimed to develop a predictive model using regression analysis for classifying an outcome for a patient as with or without diabetes based on some diagnostic measurements. 

<br>

### II. Exploratory Data Analysis 

**A. Data Collection**

The dataset used in this analysis was obtained from Kaggle which was originally from the National Institute of Diabetes and Digestive and Kidney Diseases. The dataset came from a larger database constrained to instances of patients being females, at least 21 years old and of Prima Indian heritage.  

The dataset has 768 observations and 9 indicating patient diagnostic measurements (dependent variables) including the outcome if diabetic or not (target variable). The variables are:

  * `Pregnancies`               : number of pregnancies a woman has had, including live birth, miscarriage, or stillbirth
  * `Glucose`                   : plasma glucose concentration from a 2-hour oral glucose tolerance test
  * `BloodPressure`             : diastolic blood pressure (mm Hg) 
  * `SkinThickness`             : triceps skinfold thickness (mm)
  * `Insulin`                   : two-hour serum insulin (mu U/ml)
  * `BMI`                       : body mass index
  * `DiabetesPedigreeFunction`  : risk of developing type 2 diabetes
  * `Age`                       : age
  * `Outcome`                   : is the patient with diabetes or without (1- diabetic, 0- non-diabetic)

<br>

**B. Data Exploration**

An initial look at the structure of the dataset (`diabetes_df`)shows:

```{r}
str(diabetes_df)
```
The first 10 records of `diabetes_df`:

```{r}
diabetes_df %>% head(10)
```
<br>

Summary statistics of the features:

```{r}
summary(diabetes_df)
```

Most of the features are numerical. Though expressed as the binary (0, 1), the target variable `outcome` should be categorical. Moreover, the minimum value for glucose, blood pressure, skin thickness, insulin and BMI is zero which is not medically realistic. Hence, the zero values were replaces with NA's.

```{r, echo=TRUE}
diabetes_df2 <- diabetes_df %>% 
  mutate(Outcome = as.factor(Outcome)) %>% 
  mutate(Glucose = ifelse(Glucose == 0, NA, Glucose)) %>% 
  mutate(BloodPressure = ifelse(BloodPressure == 0, NA, BloodPressure)) %>%
  mutate(SkinThickness = ifelse(SkinThickness == 0, NA, SkinThickness)) %>%
  mutate(Insulin = ifelse(Insulin == 0, NA, Insulin)) %>%
  mutate(BMI = ifelse(BMI == 0, NA, BMI)) 
```

Checking again the summary statistics:

```{r}
summary(diabetes_df2)
```
<br>

**Univariate plots**

Univariate plots are shown below.  It can be observed that most of the distributions are skewed to the right, particularly for age, insulin, pregnancies and diabetes pedigree function.  Among the patients in the dataset, 268 patients (35%) have diabetes and 500 patients (65%) are without.

```{r, fig.align='center', fig.width=15, fig.height=4}

p1_age <- ggplot(data = diabetes_df2, aes(x = Age)) +
  geom_histogram(binwidth= 3, color = "cadetblue1", fill = "cyan4") +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Age",
       y = "Count",
       title = "Age distribution")

p2_bp <- ggplot(data = diabetes_df2, aes(x = BloodPressure)) +
  geom_histogram(binwidth= 4.5, color = "cadetblue1", fill = "cyan4") +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Blood Pressure",
       y = "Count",
       title = "Blood pressure distribution")

p3_bmi <- ggplot(data = diabetes_df2, aes(x = BMI)) +
  geom_histogram(binwidth= 3, color = "cadetblue1", fill = "cyan4") +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "BMI",
       y = "Count",
       title = "BMI distribution")

ggarrange(p1_age, p2_bp, p3_bmi, ncol=3)
```
<br>

```{r, fig.align='center', fig.width=15, fig.height=4}
p4_dpf <- ggplot(data = diabetes_df2, aes(x = DiabetesPedigreeFunction)) +
  geom_histogram(binwidth= 0.1, color = "cadetblue1", fill = "cyan4") +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Diabetes Pedigree Function",
       y = "Count",
       title = "Diabetes pedigree func distribution")

p5_insulin <- ggplot(data = diabetes_df2, aes(x = Insulin)) +
  geom_histogram(binwidth= 40, color = "cadetblue1", fill = "cyan4") +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Insulin",
       y = "Count",
       title = "Insulin distribution")

p6_glucose <- ggplot(data = diabetes_df2, aes(x = Glucose)) +
  geom_histogram(binwidth= 8, color = "cadetblue1", fill = "cyan4") +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Glucose",
       y = "Count",
       title = "Glucose distribution")

ggarrange(p4_dpf, p5_insulin, p6_glucose, ncol=3)
```
<br>

```{r, fig.align='center', fig.width=15, fig.height=4}
p7_skinthickness <- ggplot(data = diabetes_df2, aes(x = SkinThickness)) +
  geom_histogram(binwidth= 5, color = "cadetblue1", fill = "cyan4") +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Skin Thickness",
       y = "Count",
       title = "Skin thickness distribution")

p8_pregnancies <- ggplot(data = diabetes_df2, aes(x = Pregnancies)) +
  geom_histogram(binwidth= 1, color = "cadetblue1", fill = "cyan4") +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Pregnancies",
       y = "Count",
       title = "Pregnancies distribution")

p9_outcome <- ggplot(data = diabetes_df2, aes(x = Outcome)) +
  geom_bar(width = 0.5, color = "cadetblue1", fill = "cyan4") +
  scale_x_discrete(labels= c("0 - non-diabetic", "1 - diabetic")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  labs(x = "Outcome",
       y = "Count",
       title = "Outcome distribution")

ggarrange(p7_skinthickness, p8_pregnancies, p9_outcome, ncol=3)
```

<br>


**Density plots based on `Outcome`**

The density plots of each predictor variable differentiating between diabetic and non-diabetic are shown below: 

```{r, fig.align='center', fig.width=15, fig.height=4}
p1_age_outcome <- ggplot(diabetes_df2, aes(x= Age, y = after_stat(density), fill = Outcome)) +
  geom_density(alpha=0.8, na.rm = TRUE) +
  scale_fill_manual(values = c("darkgoldenrod1", "cyan4")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Age",
       title = "Age density distribution")

p2_bp_outcome <- ggplot(diabetes_df2, aes(x= BloodPressure, y = after_stat(density), fill = Outcome)) +
  geom_density(alpha=0.8, na.rm = TRUE) +
  scale_fill_manual(values = c("darkgoldenrod1", "cyan4")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Blood Pressure",
       title = "Blood pressure density distribution")

p3_bmi_outcome <- ggplot(diabetes_df2, aes(x= BMI, y = after_stat(density), fill = Outcome)) +
  geom_density(alpha=0.8, na.rm = TRUE) +
  scale_fill_manual(values = c("darkgoldenrod1", "cyan4")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "BMI",
       title = "BMI density distribution")

ggarrange(p1_age_outcome, p2_bp_outcome, p3_bmi_outcome, ncol=3)

```
<br>

```{r, fig.align='center', fig.width=15, fig.height=4}

p4_dpf_outcome <- ggplot(diabetes_df2, aes(x= DiabetesPedigreeFunction, y = after_stat(density), fill = Outcome)) +
  geom_density(alpha=0.8, na.rm = TRUE) +
  scale_fill_manual(values = c("darkgoldenrod1", "cyan4")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Diabetes Pedigree Function",
       title = "Diabetes pedigree function density distribution")

p5_insulin_outcome <- ggplot(diabetes_df2, aes(x= Insulin, y = after_stat(density), fill = Outcome)) +
  geom_density(alpha=0.8, na.rm = TRUE) +
  scale_fill_manual(values = c("darkgoldenrod1", "cyan4")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Insulin",
       title = "Insulin density distribution")

p6_glucose_outcome <- ggplot(diabetes_df2, aes(x= Glucose, y = after_stat(density), fill = Outcome)) +
  geom_density(alpha=0.8, na.rm = TRUE) +
  scale_fill_manual(values = c("darkgoldenrod1", "cyan4")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Glucose",
       title = "Glucose density distribution")


ggarrange(p4_dpf_outcome, p5_insulin_outcome, p6_glucose_outcome, ncol=3)
```
<br>

```{r, fig.align='center', fig.width=15, fig.height=4}

p7_st_outcome <- ggplot(diabetes_df2, aes(x= SkinThickness, y = after_stat(density), fill = Outcome)) +
  geom_density(alpha=0.8, na.rm = TRUE) +
  scale_fill_manual(values = c("darkgoldenrod1", "cyan4")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16)) +
  theme(legend.position = "none") +
  labs(x = "Skin Thickness",
       title = "Skin thickness density distribution")

p8_pregnancies_outcome <- ggplot(diabetes_df2, aes(x= Pregnancies, y = after_stat(density), fill = Outcome)) +
  geom_density(alpha=0.8, na.rm = TRUE) +
  scale_fill_manual(values = c("darkgoldenrod1", "cyan4"), labels= c("0 - non-diabetic", "1 - diabetic")) +
  theme(plot.title = element_text(size = 15), axis.title = element_text(size = 16), axis.text = element_text(size = 16), 
        legend.title = element_text(size = 15), legend.text = element_text(size = 15),) +
  labs(x = "Pregnancies",
       title = "Pregnancies density distribution")

ggarrange(p7_st_outcome, p8_pregnancies_outcome, ncol=3, widths = c(1, 1.4, 0.6))

```
<br>

**Correlation plot**

The figure below shows the correlation among the numeric predictor variables which mostly are relatively positive. Among the variables which have quite strong correlation are: (a) skin thickness and BMI; (b) glucose and insulin; and (c) age and pregnancies (note that the dataset only includes women aged 21 and above).

```{r, include=FALSE}
diabetes_df2_num <- diabetes_df2 %>% dplyr::select(-Outcome)
```

```{r, fig.align="center", fig.width= 8, fig.height = 5}
model.matrix(~0+., data = diabetes_df2_num) %>% 
  cor(use = "pairwise.complete.obs") %>% 
  ggcorrplot(show.diag = F, type = "lower", lab = TRUE, lab_size = 3.5, colors = c("darkgoldenrod2", "white","cyan4"),
             ggtheme= ggplot2::theme_gray, tl.cex = 10)

```
<br>

**C. Data pre-processing**

Based on the summary statistics above, the dataset has a number of missing data, particularly on glucose, blood pressure, skin thickness, insulin and BMI. To address this, the missing data (zero values in the original `diabetes_df` dataset) were imputed with the mean value of the column. 

```{r, echo=TRUE}
diabetes_df3 <- diabetes_df %>% 
  mutate(Outcome = as.factor(Outcome)) %>% 
  mutate(Glucose = ifelse(Glucose == 0, mean(Glucose), Glucose)) %>% 
  mutate(BloodPressure = ifelse(BloodPressure == 0, mean(BloodPressure), BloodPressure)) %>%
  mutate(SkinThickness = ifelse(SkinThickness == 0, mean(SkinThickness), SkinThickness)) %>%
  mutate(Insulin = ifelse(Insulin == 0, mean(SkinThickness), Insulin)) %>%
  mutate(BMI = ifelse(BMI == 0, mean(BMI), BMI)) 
```

```{r, include=FALSE}
colSums(is.na(diabetes_df3))
```

<br>
<br>

### III. Evaluation of Logistic Regression Models

**A. Splitting the dataset**

The dataset was split into training data set (80%) and test data set. The training data set was used in building the logistic regression models and the test data set was used in evaluating the final model. 

```{r, echo=TRUE}
RNGkind(sample.kind = "Rounding")
set.seed(100)

train_index <-  sample(nrow(diabetes_df3), nrow(diabetes_df3)*0.8)

training_data <- diabetes_df3[train_index, ]
test_data <- diabetes_df3[-train_index, ]
```

```{r, echo=TRUE}
nrow(training_data)
nrow(test_data)
```

<br>
<br>

**B. Logistic regression models**

For this analysis, two models were evaluated:

1. Model 1 - general model including all independent variables 
2. Model 2 - stepwise model

Model results are detailed below

<br>

**Model 1: General model including all independent variables**

This model fits a logistic regression model that relates all 8 independent variables with the target `Outcome` variable.

```{r, echo=TRUE}
diabetes_logmod_01 <- glm(Outcome~ Age + BloodPressure + BMI + 
                       DiabetesPedigreeFunction + Insulin + Glucose + SkinThickness + Pregnancies, 
                       family = binomial(link = 'logit'), data = training_data)

summary(diabetes_logmod_01)
```
<br>

The results can also be presented in terms of odds ratios as in below table:

```{r}
# model summary
sjPlot::tab_model(diabetes_logmod_01, show.r2 = FALSE)
```

<br>

**Model 2 - stepwise model**

This model automatically selects a reduced number of (relevant) predictor variables for building the logistic regression model.

```{r, echo=TRUE}
diabetes_logmod_02 <- step(diabetes_logmod_01, direction = "backward", trace = 0)

summary(diabetes_logmod_02)
```

<br>

Similarly, the results can also be presented in terms of odds ratios as in below table:

```{r}
# model summary
sjPlot::tab_model(diabetes_logmod_02, show.r2 = FALSE)
```

<br>


**C. Comparing the models**

In terms of the residual deviance, Model 1 and Model 2 do not differ much. However, Model 2 has slightly lower AIC (558.94) compared to Model 1 (564.67) which makes Model 2 the better-fit model. Moreover, looking at the odds ratio tables, the variables which were excluded in Model 2 have odd ratios of 1.0 and high p-value (>0.05) meaning no association between these variables (age, blood pressure, skin thickness) and the outcome variable. Also, these variables were identified in the correlation plot as having quite strong correlations with the other variables (age, BMI). Excluding them in the model lessens multicollinearity or possibility of it which is better for a regression model.  

```{r, include= FALSE, fig.align='center', fig.width = 8, fig.height=4}
# diagnostics - residual plots
arm::binnedplot(predict(diabetes_logmod_02, type = "response"), residuals(diabetes_logmod_02, type = "response"))
```

<br>


**D. Predicting from test data**

The performance of Model 2 on new data was evaluated using the test data. Predicted values for `Outcome` were obtained as below:

```{r, echo=TRUE}
test_data$prediction <- predict(diabetes_logmod_02, newdata = test_data, type = "response")
test_data$pred_diabetic <- ifelse(test = test_data$prediction > 0.5, yes = "1", no = "0")
test_data <- test_data %>% 
  mutate(pred_diabetic = as.factor(pred_diabetic))
```

<br>

**Confusion Matrix**

A confusion matrix with the predicted and known `Outcome` values of the test data set is presented below, showing an accuracy of 0.74.

```{r}
cf <- confusionMatrix(data = test_data$pred_diabetic,
                reference = test_data$Outcome,
                positive = "1")

cf
```

```{r, fig.align='center', fig.width=4, fig.height= 4}
fourfoldplot(as.table(cf), color = c("cyan4", "goldenrod2"),main = "Confusion Matrix")
```
<br>

**ROC Curve and AUC**

The ROC (receiver operating characteristic) curve plots the sensitivity and specificity of a model. The ROC curve for Model 2 is shown below. It also indicates the AUC (area under the curve) which evaluates how well this logistic regression model classifies the `Outcomes` outcomes based on the set cutoff (0.5). A high AUC means the model is better at predicting non-diabetic (0 classes as 0) and diabetic (1 classes as 1). Model 2 resulted to a moderately good AUC of 0.788.

```{r, fig.align='center', fig.width=5, fig.height= 5}
pROC::plot.roc(test_data$Outcome, as.numeric(test_data$prediction), print.auc = TRUE)
```

```{r}
pROC::auc(test_data$Outcome, as.numeric(test_data$prediction))
```

<br>
<br>

### IV. Conclusion

In this project, a logistic regression model was developed to predict if a patient is diabetic or not. Among the two models evaluated, Model 2 which included reduced predictor variables has the better fit in classifying the `Outcome`. 

From Model 2 results, holding other variables at fixed value: (a) a unit increase in glucose concentration increases the odds of the patient being diabetic by 4%; (b) a new pregnancy increases the odds of the patient being diabetic by 15%; (c) an increase in BMI increases the odds of a patient being diabetic by 10%; and a unit increase in the diabetes pedigree function score increases the odds of a patient being diabetic by a factor of 4.14 (or 4 times). Insulin has an odds ration of 1.0 which means no association with the outcome.

Based on the obtained AUC, the model has 79%% chance of distinguishing a patient as diabetic (1 class) or not (0 class). The dataset seemed not large enough and hence, additional observations (data) might improve the model. 











